\section{Control Structures}
\markright{\arabic{section}. Control Structures}
\subsection{Conditionals}

Although {\bf and, or} and {\bf cond} are advised to be macros by Common Lisp,
they are implemented as special forms in EusLisp to improve
the interpreting performance.

\begin{refdesc}
\specialdesc{and}{\{form\}*}{
{\em Form}s are evaluated from left to right until NIL appears.
If all forms are evaluated to non-NIL, the last value is returned.}

\specialdesc{or}{\{form\}*}{
{\em Form}s are evaluated from left to right until non-NIL appears,
and the value is returned. If all forms are evaluated to NIL, 
NIL is returned.}

\specialdesc{if}{test then [else]}{
{\bf if} can only have single {\it then} and {\it else} forms.
To allow multiple {\em then} or {\em else} forms,
they must be grouped by {\bf progn}.}

\macrodesc{when}{test forms}{
Unlike {\bf if},
{\bf when} and {\bf unless} allow you to write multiple {\em forms}
which are executed when {\em test} holds ({\bf when}) or
does not {\em unless}.
On the other hand, these macros
cannot have the {\em else} forms.}

\macrodesc{unless}{test forms}{
is equivalent to {\tt(when (not {\em test}) . {\em forms})}.}

\specialdesc{cond}{{(test \{form\}*)}*}{
Arbitrary number of  cond-clauses can follow {\bf cond}.
In each clause, the first form, that is {\em test}, is evaluated.
If it is non-nil, the rest of the forms in that clause are evaluated sequentially,
and the last value is returned. 
If no forms are given after the {\em test}, the value of the {\em test} is returned.
When the {\em test} fails, next clause is tried until a {\em test} which is evaluated
to non-nil is found or all clauses are exhausted.
In the latter case, {\bf cond} returns NIL.}

\macrodesc{case}{key \{(\{label $|$ (\{lab\}*) \{form\}*)\}*}{
For the clause whose {\em label} matches with {\em key},
{\em form}s are evaluated and the last value is returned.
Equality between {\em key} and {\em label} is tested with {\bf eq}
or {\bf memq}, not with {\bf equal}.}

\end{refdesc}

\subsection{Sequencing and Lets}

\begin{refdesc}
\funcdesc{prog1}{form1 \&rest forms}{
{\em form1} and {\em forms} are evaluated sequentially, 
and the value returned by {\em form1} is returned as the value of {\bf prog1}.}

\specialdesc{progn}{\{form\}*}{
{\em Form}s are evaluated sequentially, and the value of the rightmost form
is returned.
{\bf Progn} is a special form because it has a special meaning when it
appeared at top level in a file.
When such a form is compiled, all inner forms are regarded as they appear
at top level.
This is useful for a macro which expands to a series of
{\bf defun}s or {\bf defmethod}s, which must appear at top level.}

\macrodesc{setf}{\{access-form value\}*}{
assigns {\em value} to a generalized-variable {\em access-form}.}

\specialdesc{let}{(\{var $|$ (var [value])\}*) \{declare\}* \{form\}*}{
introduces local variables.
All {\em value}s are evaluated and assigned to {\em var}s in parallel, i.e.,
{\tt (let ((a 1)) (let ((a (1+ a)) (b a)) (list a b)))} produces
(2 1).}

\specialdesc{let*}{(\{var $|$ (var [value])\}*) \{declare\}* \{form\}*}{
introduces local variables.
All {\em value}s are evaluated sequentially, and assigned to {\em var}s
i.e.,
{\tt (let ((a 1)) (let* ((a (1+ a)) (b a)) (list a b)))} produces
(2 2).}
\end{refdesc}

\subsection{Local Functions}
\begin{refdesc}
\specialdesc{flet}{ (\{(fname lambda-list . body)\}*) \{form\}*}{
defines local functions.}

\specialdesc{labels}{ (\{(fname lambda-list . body)\}*) \{form\}*}{
defines locally scoped functions. 
The difference between {\em flet} and {\em labels} is,
the local functions defined by {\em flet} cannot reference
each other or recursively, whereas {\em labels} allows such mutual references.}
\end{refdesc}

\subsection{Blocks and Exits}
\begin{refdesc}
\specialdesc{block}{tag \{form\}*}{
makes a lexical block from which you can exit by {\bf return-from}.
{\em Tag} is lexically scoped and is not evaluated.}

\specialdesc{return-from}{tag value}{
exits the block labeled by {\em tag}.
{\bf return-from} can be used to exit from a function or a method which
automatically establishes block labeled by its function or method name
surrounding the entire body.}

\macrodesc{return}{value}{
{\tt  (return x)} is equivalent to {\tt (return-from nil x)}. This is
convenient to use in conjunction with {\bf loop, while, do, dolist,}
and {\bf dotimes} which implicitly establish blocks labeled NIL.}

\specialdesc{catch}{ tag \{form\}*}{
establishes a dynamic block from which you can exit and return a value
by {\bf throw}. {\em Tag} is evaluated.
The list of all visible catch tags can be obtained by {\tt
sys:list-all-catchers}.}

\specialdesc{throw}{tag value}{
exits and returns {\em value} from a catch block.
{\em tag} and {\em value} are evaluated.}

\specialdesc{unwind-protect}{protected-form \{cleanup-form\}*}{
After the evaluation of {\em protected-form} finishes,
{\em cleanup-form} is evaluated.
You may make a block or a catch block outside the {\tt unwind-protect}.
Even {\bf return-from} or {\bf throw} is executed in {\em protected-form}
to escape from such blocks, {\em cleanup-form} is assured to be evaluated.
Also, if you had an error while executing {\em protected-form},
{\em cleanup-form} would always be executed by {\em reset}.}

\end{refdesc}

\subsection{Iteration}

\begin{refdesc}

\specialdesc{while}{test \{form\}*}{
While {\em test} is evaluated to non-nil,
{\em form}s are evaluated repeatedly.
{\bf While} special form automatically establishes a block by name of nil
around {\em form}s, and {\bf return} can be used to exit from the loop.
To jump to next iteration, you can use following syntax with {\bf tagbody} and {\bf go} described below:}

\begin{verbatim}
(setq cnt 0)
(while
  (< cnt 10)
  (tagbody while-top
    (incf cnt)
    (when (eq (mod cnt 3) 0)
      (go while-top))  ;; jump to next iteraction
    (print cnt)
    )) ;; 1, 2, 4, 5, 7, 8, 10
\end{verbatim}

\specialdesc{tagbody}{\{tag $|$ statement\}*}{
{\em tag}s are labels for {\bf go}.
You can use {\bf go} only in {\bf tagbody}.}

\specialdesc{go}{tag}{
transfers control to the form just after {\em tag}
which appears in a lexically scoped {\bf tagbody}.
{\bf Go} to the tag in a different {\bf tagbody}
across the lexical scope is inhibited.}

\macrodesc{prog}{(\{var $|$ (var [init])\}*) \{tag $|$ statement\}*}{
{\bf prog} is a macro, which expands as follows:
\ptext{
(block nil 
    (let {\em var}
	(tagbody
		{\em tag $|$ statement})))}
}

\macrodesc{do}{(\{(var init [next])\}*) (endtest [result])\{declare\} \{form\}
*}
{{\em var}s are local variables.
To each {\em var}, {\em init} is evaluated in parallel and assigned.
Next, {\em endtest} is evaluated and if it is true, {\bf do} returns
{\em result} (defaulted to NIL).
If {\em endtest} returns NIL, each {\em form} is evaluated sequentially.
After the evaluation of forms, {\em next} is evaluated and the value is
reassigned to each {\em var}, and the next iteration starts.}

\macrodesc{do*}{(\{var init [next]\}*) (endtest [result])\{declare\} \{form\}*}
{{\bf do*} is same as {\bf do} except that the evaluation of {\em init}
and {\em next}, and their assignment to {\em var} occur sequentially.}

\macrodesc{dotimes}{(var count [result]) \{forms\}*}{
evaluates {\em forms} {\em count} times.
{\em count} is evaluated only once.
In each evaluation, {\em var} increments from integer zero to
{\em count} minus one.}

\macrodesc{dolist}{(var list [result]) \{forms\}*}{
Each element of {\em list} is  sequentially bound to {\em var},
and {\em forms} are evaluated for each binding.
{\bf Dolist} runs faster than other iteration constructs
such as {\bf mapcar} and recursive functions,
since {\bf dolist} does not have to create a function closure or to apply it,
and no new parameter binding is needed.}

\macrodesc{until}{condition \{forms\}*}{
evaluates forms until {\em condition} holds.}

\macrodesc{loop}{\{forms\}*}{
evaluates {\em forms} forever.
To terminate execution, {\bf return-from, throw} or {\bf go} needed to be
evaluated in {\em forms}.}

\subsection{Predicates}

{\bf Typep} and {\bf subtypep} of Common Lisp are not provided, and should be
simulated by {\bf subclassp} and {\bf derivedp}.

\begin{refdesc}

\funcdesc{eq}{obj1 obj2}{returns T,
if {\em obj1} and {\em obj2} are pointers to the same object,
or the same numbers.
Examples: {\tt (eq 'a 'a)} is T, {\tt (eq 1 1)} is T,
{\tt (eq 1. 1.0)} is nil, {\tt (eq "a" "a")} is nil.}
\funcdesc{eql}{obj1 obj2}{
{\bf Eq} and {\bf eql} are identical since all the numbers in EusLisp are represented as
immediate values.}
\funcdesc{equal}{obj1 obj2}{
Checks the equality of any structured objects, such as strings, vectors or
matrices, as long as they do not have recursive references.
If there is recursive reference in {\em obj1} or {\em obj2},
{\bf equal} loops infinitely.}
\funcdesc{superequal}{obj1 obj2}{
Slow but robust {\bf equal}, since {\bf superequal} checks circular reference.}

\funcdesc{null}{object}{T if {\em object} is nil.
Equivalent to (eq {\em object} nil).}
\funcdesc{not}{object}{
{\bf not} is identical to {\bf null}.}
\funcdesc{atom}{object}{
returns NIL only if object is a cons.
{\tt (atom nil) = (atom '()) = T)}.
Note that {\bf atom} returns T for vectors, strings, read-table, hash-table, 
etc., no matter what complex objects they are.}
\funcdesc{every}{pred \&rest args}{
returns T if all {\em args} return T for {\em pred}.
{\bf Every} is used to test whether {\em pred} holds for every {\em args}.}
\funcdesc{some}{pred \&rest args}{
returns T if at least one of {\em args} return T for {\em pred}.
{\bf Some} is used to test whether {\em pred} holds for any of {\em args}.}
\end{refdesc}
\funcdesc{functionp}{object}{
T if {\em object} is a function object that can be given to
{\bf apply} and {\bf funcall}.
Note that macros cannot be {\em apply}'ed or {\em funcall}'ed.
{\bf Functionp} returns T, if {\em object} is
either a compiled-code with type=0, a symbol that has function definition,
a lambda-form, or a lambda-closure.
Examples: (functionp 'car) = T, (functionp 'do) = NIL} 
\funcdesc{compiled-function-p}{object}{
T if {\em object} is an instance of compiled-code.
In order to know the compiled-code is a function or a macro,
send {\tt :type} message to the object, and {\tt function} or {\tt macro}
is returned.}

\end{refdesc}

\newpage
