\section{Date and Time}

The time class defines both calendar time and time period.

\begin{refdesc}

\classdesc{time}{propertied-object}
{(micro second minute hour day month weekday year \\
\> timezone dst seconds)}{
defines time objects.
}

\methoddesc{:now}{}{
\verb~(instance time :now)~
creates a time object for the current time.}

\methoddesc{:init}{\&optional sec micro dst tzone}{
creates a time object which represents {\it sec} second
after January 1, 1970.}

\methoddesc{:make}{\&key (year 1970) (month 0) (day 1)
	(weekday 4) (hour 0) (minute 0) (second 0)  (micro 0)
	(timezone (* -9 3600)))}{
creates a time object which is represented
by a calendar notation.
The timezone is defaulted to JST.}

\methoddesc{:year}{}{returns the year component of the
time object.  Note that the year is represented in a full (four)
digit notation, not the least two digits.}

\methoddesc{:month}{}{returns the month component of the time object.
Note that the month begins from 0 for January.}

\methoddesc{:day}{}{returns the day component of the time object.
Note that the day begins from 1 for the first of a month.}

\methoddesc{:weekday}{}{returns the weekday component of the time object.
Note that the weekday begins from 0 for Sunday.}

\methoddesc{:hour}{}{returns the hour component of the time object
in 24-hour representation.
Note that the hour ranges from 0 to 23.}

\methoddesc{:minute}{}{returns the minute component of the time object.
Note that the hour ranges from 0 to 59.}

\methoddesc{:second}{}{returns the second component of the time object.
Note that the hour ranges from 0 to 59.}

\methoddesc{:seconds}{}{returns the seconds component of the time object.
Seconds represents time after the origin of the unix time, i.e.,
the midnight of January 1, 1970.}

\methoddesc{:year-day}{}{returns the number of days after the
beginning of the year. For example, year-day of a time object representing
February 2nd is 32.}

\methoddesc{:difference}{atime}{returns a new time object
representing the time difference of self from atime.}

\methoddesc{:add}{atime}{returns a new time object
representing the added time of self and atime.}

\end{refdesc}




 
