\macrodesc{load-foreign}{objfile \&key symbol-input symbol-output (symbol-file objfile) ld-option)}
{loads an object module written other than EusLisp.
A compiled code object is returned.
This result is necessary to make function entries in the module by
{\bf defforeign}.
Libraries can be specified in {\em ld-option}.
However, the symbols defined in these libraries cannot be captured
in the default symbol-output file.
In order to allow EusLisp to call functions defined in the libraries,
{\em symbol-output} and {\em symbol-file} must be given explicitly.
(These arguments are not needed if you are not going to call the library
functions directly from EusLisp and if you are  referring them only from
{\em objfile})
{\bf load-foreign} links {\em objfile} with libraries specified and global
symbols in EusLisp which is in core, and writes the linked object in
{\em symbol-output}.
Then, symbols in {\em symbol-file} are searched and listed in the
foreign-module. Since {\em symbol-file} is defaulted to be {\em objfile},
only the symbols defined in {\em objfile} are  recognized if {\em symbol-file}
is not given. To find all the global entries both in {\em objfile} and
libraries, the linked (merged) symbol table resulted from the first link
process of load-foreign must be examined.
For this reason, an identical file name must be given to {\em symbol-output}
and {\em symbol-file}.

As shown below, the intermediate symbol file can be removed
by {\bf unix:unlink}.
However, if you are loading more than one foreign modules both of which 
refer to the same library, and if you want to avoid loading
the library duplicatedly, you have to use {\em symbol-input} argument.
Suppose you have loaded all the functions in "linpack.a" in the above
example and you are going to load another file "linapp.o" that calls functions
in "linpack.a".
The following call of load-foreign should be issued before you
unlink "euslinpack".
{\tt (load-foreign "linapp.o" :symbol-input "euslinpack")}
}

\begin{verbatim}
(setq linpack-module
    (load-foreign "/usr/local/eus/clib/linpackref.o"
        :ld-option  "-L/usr/local/lib -llinpack -lF77 -lm -lc"
        :symbol-output "euslinpack"
        :symbol-file "euslinpack"
        ))
(unix:unlink "euslinpack")
\end{verbatim}

